Vagueness \& Rationality

### 2018-10-27 manuscript/e1.tex
removed undirected quote marks and replaced with aostrophe-style quotes

\begin{itemize}
	\item In the vague numerical condition the instruction was ``Choose a square with about 10 dots" -- none of the arrays contained exactly 10 dots, but 10 is closer to 6 than it is to 15, making 6 a better response to that instruction, 15 a borderline response, and 24 an extreme response. 
	\item In the vague verbal condition we used ``Choose a square with few dots". We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
	\item In the crisp numerical condition we used ``Choose the square with 6 dots". The smaller flanking array always contained exactly the specified number of dots. We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
	\item For crisp verbal, we used ``Choose the square with the fewest dots". We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
\end{itemize}

On each trial, first the referring expression that constituted the instruction for that trial was displayed (e.g., ``Choose a square with about 10 dots"). Participants then pressed a key to indicate that they had read the instruction. The instruction remained on screen, and after 1000 ms, the arrays appeared. An example stimulus is given in Figure \ref{Experiment1and2examplestimulus}. Response time was measured from the presentation of the arrays until the keypress indicating the participant's choice. The trial would timeout after 60 seconds if there was no response. In this experiment, no feedback was given. This was because, in the vague conditions, we did not regard any response as ``correct" or ``incorrect", but instead as ``expected response"; ``borderline response"; and ``extreme response", and we did not want to draw participants' attention to this distinction explicitly. Which choice the participant made was recorded for analysis.

\begin{table}[htbp]
\centering
\caption{Experiment 1 instructions arranged by condition. The instructions given in the table started with ``Choose \ldots"}
\label{instructionsC-exp-1}
\begin{tabular}{cccll}
\hline\noalign{\smallskip}
Item & Quantity & Number & Crisp & Vague\\
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 & Small & Numeric & the square with 6 dots & a square with about 10 dots\\
06:15:24 & Small & Verbal & the square with the fewest dots & a square with few dots\\
06:15:24 & Large & Numeric & the square with 24 dots & a square with about 20 dots\\
06:15:24 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
16:25:34 & Small & Numeric & the square with 16 dots & a square with about 20 dots\\
16:25:34 & Small & Verbal & the square with the fewest dots & a square with few dots\\
16:25:34 & Large & Numeric & the square with 34 dots & a square with about 30 dots\\
16:25:34 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
26:35:44 & Small & Numeric & the square with 26 dots & a square with about 30 dots\\
26:35:44 & Small & Verbal & the square with the fewest dots & a square with few dots\\
26:35:44 & Large & Numeric & the square with 44 dots & a square with about 40 dots\\
26:35:44 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
36:45:54 & Small & Numeric & the square with 36 dots & a square with about 40 dots\\
36:45:54 & Small & Verbal & the square with the fewest dots & a square with few dots\\
36:45:54 & Large & Numeric & the square with 54 dots & a square with about 50 dots\\
36:45:54 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

\begin{figure}[htbp]
\centering
\includegraphics[width=.75\textwidth]{figures/Ce1-example-screenshot}
\caption{Experiment 1 and 2 example stimulus}
\label{Experiment1and2examplestimulus}
\end{figure}

@@

However, given that the response time plot in Figure \ref{resultsC-exp-1-RT} shows that responses to 6:15:24 in the "crisp numeric" instructions condition were extremely fast relative to the "vague numeric" instructions to 6:15:24, the effects in the model of the full dataset could be driven by this difference. A clearer picture of the effects of interest might be obtained by removing the 6:15:24 level of Item from the data set, and fitting the model to this restricted data. Doing this did not affect the direction of the effects in the full dataset, but whereas the effects were significant in the full dataset, they were not significant in the restricted dataset. Full details of the analysis of the restricted dataset are available in the online materials.

@@

However, once again there is a possibly confounding factor. Observe that, in Experiment 1, instruction format (i.e., the difference between numeric and verbal) went hand in hand with might be called the (human) "selection algorithm": To see this, consider the task of selecting the dot array that contains "few dots":" to do this, it suffices to \emph{compare} the three arrays and select the one that contains the fewest elements.  To select the dot array that contains "16 dots" seems to require the participant to estimate, and then \emph{match}, the cardinality of (at least) one dot array to 16, a process which could plausibly take longer, independently of vagueness. Therefore, our results so far permit the interpretation that what made the instructions in the verbal condition fast is not the fact that they were worded verbally, but that they allowed participants to use \emph{comparison} rather than having to resort to \emph{matching}.

>>>>

\begin{itemize}
	\item In the vague numerical condition the instruction was ``Choose a square with about 10 dots'' -- none of the arrays contained exactly 10 dots, but 10 is closer to 6 than it is to 15, making 6 a better response to that instruction, 15 a borderline response, and 24 an extreme response.
	\item In the vague verbal condition we used ``Choose a square with few dot''. We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
	\item In the crisp numerical condition we used ``Choose the square with 6 dots''. The smaller flanking array always contained exactly the specified number of dots. We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
	\item For crisp verbal, we used ``Choose the square with the fewest dots''. We considered this to be equivalent in terms of which responses were expected (6), borderline (15) and extreme (24).
\end{itemize}

On each trial, first the referring expression that constituted the instruction for that trial was displayed (e.g., ``Choose a square with about 10 dots''). Participants then pressed a key to indicate that they had read the instruction. The instruction remained on screen, and after 1000 ms, the arrays appeared. An example stimulus is given in Figure \ref{Experiment1and2examplestimulus}. Response time was measured from the presentation of the arrays until the keypress indicating the participant's choice. The trial would timeout after 60 seconds if there was no response. In this experiment, no feedback was given. This was because, in the vague conditions, we did not regard any response as ``correct'' or ``incorrect'', but instead as ``expected response''; ``borderline response''; and ``extreme response'', and we did not want to draw participants' attention to this distinction explicitly. Which choice the participant made was recorded for analysis.

\begin{table}[htbp]
\centering
\caption{Experiment 1 instructions arranged by condition. The instructions given in the table started with ``Choose \ldots''}
\label{instructionsC-exp-1}
\begin{tabular}{cccll}
\hline\noalign{\smallskip}
Item & Quantity & Number & Crisp & Vague\\
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 & Small & Numeric & the square with 6 dots & a square with about 10 dots\\
06:15:24 & Small & Verbal & the square with the fewest dots & a square with few dots\\
06:15:24 & Large & Numeric & the square with 24 dots & a square with about 20 dots\\
06:15:24 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
16:25:34 & Small & Numeric & the square with 16 dots & a square with about 20 dots\\
16:25:34 & Small & Verbal & the square with the fewest dots & a square with few dots\\
16:25:34 & Large & Numeric & the square with 34 dots & a square with about 30 dots\\
16:25:34 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
26:35:44 & Small & Numeric & the square with 26 dots & a square with about 30 dots\\
26:35:44 & Small & Verbal & the square with the fewest dots & a square with few dots\\
26:35:44 & Large & Numeric & the square with 44 dots & a square with about 40 dots\\
26:35:44 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline\noalign{\smallskip}
36:45:54 & Small & Numeric & the square with 36 dots & a square with about 40 dots\\
36:45:54 & Small & Verbal & the square with the fewest dots & a square with few dots\\
36:45:54 & Large & Numeric & the square with 54 dots & a square with about 50 dots\\
36:45:54 & Large & Verbal & the square with the most dots & a square with many dots\\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

\begin{figure}[htbp]
\centering
\includegraphics[width=.75\textwidth]{figures/Ce1-example-screenshot-border}
\caption{Experiment 1 and 2 example stimulus}
\label{Experiment1and2examplestimulus}
\end{figure}

@@

However, given that the response time plot in Figure \ref{resultsC-exp-1-RT} shows that responses to 6:15:24 in the `crisp numeric' instructions condition were extremely fast relative to the `vague numeric' instructions to 6:15:24, the effects in the model of the full dataset could be driven by this difference. A clearer picture of the effects of interest might be obtained by removing the 6:15:24 level of Item from the data set, and fitting the model to this restricted data. Doing this did not affect the direction of the effects in the full dataset, but whereas the effects were significant in the full dataset, they were not significant in the restricted dataset. Full details of the analysis of the restricted dataset are available in the online materials.

@@

However, once again there is a possibly confounding factor. Observe that, in Experiment 1, instruction format (i.e., the difference between numeric and verbal) went hand in hand with might be called the (human) `selection algorithm': To see this, consider the task of selecting the dot array that contains `few dots': to do this, it suffices to \emph{compare} the three arrays and select the one that contains the fewest elements.  To select the dot array that contains `16 dots' seems to require the participant to estimate, and then \emph{match}, the cardinality of (at least) one dot array to 16, a process which could plausibly take longer, independently of vagueness. Therefore, our results so far permit the interpretation that what made the instructions in the verbal condition fast is not the fact that they were worded verbally, but that they allowed participants to use \emph{comparison} rather than having to resort to \emph{matching}.

### 2018-10-27 manuscript/e3.tex
removed undirected quote marks and replaced with aostrophe-style quotes

\begin{figure}[htbp]
\centering
\includegraphics[width=.5\textwidth]{figures/Ee3-flow.jpg}
\caption{Experiment 3 example stimulus, showing the sequence of presentation, with the instruction screen at the top, followed by the presentation of the "target" array, followed by the stimulus that participants responded to.}
\label{Experiment3examplestimulus}
\end{figure}

\begin{table}
\centering
\caption{Experiment 3 instructions for the smallest array, arranged by condition. The instructions given in the table started with ``Choose a square with \ldots"}
\label{instructionsE-exp-3}
\resizebox{\textwidth}{!}{\begin{tabular}{ccccl}
\hline\noalign{\smallskip}
Item & Target & Selection & Vagueness & Instruction \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 &   6 & Matching 	& Crisp 	& the same number of dots as the target 	    \\ 
06:15:24 &  10 & Matching	& Vague		& about the same number of dots as the target	\\
06:15:24 &  20 & Comparison & Crisp 	& fewer dots than the target             	    \\ 
06:15:24 &  20 & Comparison	& Vague		& far fewer dots than the target				\\
\noalign{\smallskip}\hline
\end{tabular}}
\end{table}

>>>>

\begin{figure}[htbp]
\centering
\includegraphics[width=.5\textwidth]{figures/Ee3-flow.jpg}
\caption{Experiment 3 example stimulus, showing the sequence of presentation, with the instruction screen at the top, followed by the presentation of the ``target'' array, followed by the stimulus that participants responded to.}
\label{Experiment3examplestimulus}
\end{figure}

\begin{table}
\centering
\caption{Experiment 3 instructions for the smallest array, arranged by condition. The instructions given in the table started with ``Choose a square with \ldots''}
\label{instructionsE-exp-3}
\resizebox{\textwidth}{!}{\begin{tabular}{ccccl}
\hline\noalign{\smallskip}
Item & Target & Selection & Vagueness & Instruction \\ 
\noalign{\smallskip}\hline\noalign{\smallskip}
06:15:24 &   6 & Matching 	& Crisp 	& the same number of dots as the target 	    \\ 
06:15:24 &  10 & Matching	& Vague		& about the same number of dots as the target	\\
06:15:24 &  20 & Comparison & Crisp 	& fewer dots than the target             	    \\ 
06:15:24 &  20 & Comparison	& Vague		& far fewer dots than the target				\\
\noalign{\smallskip}\hline
\end{tabular}}
\end{table}

### 2018-10-27 manuscript/our-approach-to-the-problem.tex
removed undirected quote marks and replaced with aostrophe-style quotes

In Experiment 1, reported below, we also addressed another issue with the second pilot experiment: participants chose one of two dot arrays - therefore the `vague' quantifiers (few and many) uniquely identified one square. Recall our definition of vague  -- ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise''. The quantifiers few and many might not have realised their potential for vagueness in the context of a choice between (only) two alternatives - where there is no borderline alternative that could result in few and many being "not well-defined". Furthermore using definite articles in the instructions may have reinforced the participant's impression that only one choice counted as correct, and this impression could have been reinforced by the use of error feedback.

>>>>

In Experiment 1, reported below, we also addressed another issue with the second pilot experiment: participants chose one of two dot arrays - therefore the `vague' quantifiers (few and many) uniquely identified one square. Recall our definition of vague  -- ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise''. The quantifiers few and many might not have realised their potential for vagueness in the context of a choice between (only) two alternatives - where there is no borderline alternative that could result in few and many being ``not well-defined''. Furthermore using definite articles in the instructions may have reinforced the participant's impression that only one choice counted as correct, and this impression could have been reinforced by the use of error feedback.

### 2018-10-27 manuscript/related-work.tex
removed undirected quote marks and replaced with aostrophe-style quotes

A number of answers to Lipman's question have been proposed, but as we have argued elsewhere, it appears that some of these do not to hold up to scrutiny \citep{van2009utility,vanDeemterBook}. For example,
%
\begin{itemize}
\item It has been proposed that the existence of vague words makes a language more efficient \citep{barwiseperry}. The idea is that vagueness allows the use of one and the same word (e.g., ``big'') in different situations, which require different standards. For instance, the vague word ``big'' can denote a mouse a well as an elephant. The idea is plausible, but the problem is that ``big" is efficient not because it is vague (i.e., not because it allows borderline cases) but because it is context dependent. To see this, note for example that superlatives are not vague (i.e., they do not permit borderline cases), but they {\em are} context dependent: ``the biggest $x$" ascribes different sizes to the referent depending on $x$. It is for this reason that we can use the word ``biggest''  to talk about the biggest mouse, the biggest elephant, and so on. Context-dependence, however, does not imply vagueness.
%
\item Vague words are capable of combining a statement of quantity with an evaluative statement \citep{veltman}. The idea is that when we say ``The patient has a fever'', we do not merely assert that her temperature lies above a certain value, we also imply that the deviation is clinically significant (i.e., something is not right with the patient). The problem with this plausible idea is that evaluation does not imply vagueness. For example, the medical term ``obese'' is evaluative (i.e., it is not healthy to be obese), yet its standard medical definition in terms of Body Mass Index is perfectly crisp (i.e., without borderline cases), with all and only BMI values above 30 counting as obese.
\end{itemize}
%
Analogous observations can be made about a number of other purported benefits of vagueness; one of the few hypotheses left standing at the moment is the {\em cost reduction} hypothesis, which we mentioned in the previous section. The findings on which we are reporting in the present paper will follow a familiar pattern: we will explore the cost reduction hypothesis, only to conclude that it may not stand up to scrutiny. Our approach, this time, will be experimental: we conduct controlled experiments to investigate the effect that vague expressions have on a hearer.

>>>>

A number of answers to Lipman's question have been proposed, but as we have argued elsewhere, it appears that some of these do not to hold up to scrutiny \citep{van2009utility,vanDeemterBook}. For example,
%
\begin{itemize}
\item It has been proposed that the existence of vague words makes a language more efficient \citep{barwiseperry}. The idea is that vagueness allows the use of one and the same word (e.g., ``big'') in different situations, which require different standards. For instance, the vague word ``big'' can denote a mouse a well as an elephant. The idea is plausible, but the problem is that ``big'' is efficient not because it is vague (i.e., not because it allows borderline cases) but because it is context dependent. To see this, note for example that superlatives are not vague (i.e., they do not permit borderline cases), but they {\em are} context dependent: ``the biggest $x$'' ascribes different sizes to the referent depending on $x$. It is for this reason that we can use the word ``biggest''  to talk about the biggest mouse, the biggest elephant, and so on. Context-dependence, however, does not imply vagueness.
%
\item Vague words are capable of combining a statement of quantity with an evaluative statement \citep{veltman}. The idea is that when we say ``The patient has a fever'', we do not merely assert that her temperature lies above a certain value, we also imply that the deviation is clinically significant (i.e., something is not right with the patient). The problem with this plausible idea is that evaluation does not imply vagueness. For example, the medical term ``obese'' is evaluative (i.e., it is not healthy to be obese), yet its standard medical definition in terms of Body Mass Index is perfectly crisp (i.e., without borderline cases), with all and only BMI values above 30 counting as obese.
\end{itemize}
%
Analogous observations can be made about a number of other purported benefits of vagueness; one of the few hypotheses left standing at the moment is the {\em cost reduction} hypothesis, which we mentioned in the previous section. The findings on which we are reporting in the present paper will follow a familiar pattern: we will explore the cost reduction hypothesis, only to conclude that it may not stand up to scrutiny. Our approach, this time, will be experimental: we conduct controlled experiments to investigate the effect that vague expressions have on a hearer.
